Revisiting the Relationship between Short-Chain Fatty Acids, the Microbiota, and Colorectal Tumors

### 2018-07-27 submission/manuscript.Rmd
second paragraph half changed
update discussion and conclusion sections

Much of the previous research on SCFA benefit to human CRC has been illustrated in model systems [@test_OKeefe2016]. Many SCFAs are produced through the breakdown of fiber [@Holscher2017] and a recent study in mice found that fiber supplementation increased SCFA concentrations and decreased tumor burden [@Bishehsari2018]. Additionally, SCFAs such as butyrate can inhibit tumor growth in *in vitro* experiments [@Encarnao2018]. Yet, observations in humans has been mixed. Previous case/control studies found associations with lower SCFA concentrations in individuals with carcinoma tumors [@Ohigashi2013]. However, individual randomized controlled trials and a recent meta-analysis on fiber supplementation to prevent tumor recurrence has found no benefit [@Schatzkin2000; @Yao2017]. Our results align with what has been reported in randomized-controlled trials, that SCFAs do not provide general protection against colorectal tumors. It is possible though that there are specific instances where SCFAs may be beneficial.

One limitation of current research into the effect of SCFAs in CRC has been that all tumors are treated as the same type. However, there are known differences in the types of mutations that occur [@Fearon1990] and treating all tumors as equal may actually hide any benefit that could be found in specific subsets of individuals. Similar to the idea of specific immunotherapy for specific tumors [@Thomas2016], SCFAs may have beneficial effects for specific types of colorectal tumors. Future research will need to test if this is a valid hypothesis. Another limitation is that a fecal sample may not be an ideal type of biospecimen and that the effect SCFAs have on tumorigenesis is only detected in the colon. However, most *in vivo* studies as well as human studies have used fecal material in their analysis [@Ohigashi2013; @Bishehsari2018]. Additionally, studies that measure SCFA changes after fiber supplementation use fecal material to track these responses [@scfa_measures_venkataraman2016]. Although there are limitations with the current research on SCFAs and colorectal tumors, our observations along with the randomized controlled trials on fiber supplementation in tumor recurrence [@Yao2017] provide evidence that these specific metabolites may not be protective. Yet, taxa that are associated with SCFA production are consistently higher in indivdiuals without colon tumors than patients with carcinomas [@Shah2017; @normalization_Sze2017; @meta_analysis_crc_Sze2018].

The potential protection against colorectal cancers may not be from SCFAs even though taxa associated with their production are higher in individuals without tumors versus patients with carcinomas [@Shah2017; @normalization_Sze2017; @meta_analysis_crc_Sze2018]. Protection could be via a different pathway and by extension other metabolites that have not been extensively studied. Alternatively, protection may not occur via a metabolite but instead through niche exclusion of mouth-associated microbes (e.g. *Fusobacterium*, *Porphyromonas*, *Parvimonas*, *Peptostreptococcus* [@Zeller2014; @Baxter2016; @Kostic2011]). The idea of niche exclusion is similar to how the community protects against *Clostridium difficile* infection [@Theriot2015] with chronic inflammation replacing the role of antibiotics. Although we did not find lower concentrations of SCFAs associated with colorectal tumors, we think that there are many exciting new avenues to explore because of these results.

@@

Our observations found no difference in SCFA concentration, their utility as a classification tool, or for genes of enzymes involved in SCFA synthesis between individuals without colon tumors and patients with either adenoma or carcinoma tumors. Although these results are different than other reports in the literature, they do align with the randomized controlled trials that have tested fiber use in preventing colorectal tumor recurrence. Overall, these results suggest that the SCFAs typically produced by resident microbes do not protect against tumor. By focusing on other types mechanisms, the identification of more promising therapeutic options for use in treating colorectal cancer may be found.

>>>>

Although SCFAs have been shown to be anti-tumorigenic, most of these studies have been performed in model systems [@test_OKeefe2016; @Encarnao2018]. Additionally, many of the *in vivo* studies investigate proxies such as fiber supplementation rather than SCFAs directly [@Holscher2017]. Although it is well known that breakdown products from gut bacteria results in SCFA production, fiber effects on tumorigenesis may be through other mechanisms in these *in vivo* model systems. In addition to using fiber supplementation as a proxy for SCFA treatment, observations on the benefit of SCFAs in preventing tumorigenesis has been mixed. In previous case/control studies lower SCFA concentrations between patients with carcinomas and those without carcinomas was observed [@Ohigashi2013]. Yet, this is in contrast to multiple randomized-controlled trials that have found no difference between patients who do and do not get fiber supplementation to try and prevent tumor recurrence [@Schatzkin2000; @Yao2017]. This could be because fiber has other actions in humans besides providing an energy source for bacteria to breakdown to make SCFAs. It could also be due to the fact that SCFA concentrations and responses to fiber can vary quite a bit between healthy individuals [@scfa_measures_venkataraman2016]. This information taken together with our observations would suggest that either indviduals who do not respond to fiber supplementation would need to acquire these bacteria to achieve a benefit or that SCFAs provide little to no benefit as an anti-tumorigenic compound in colorectal cancer.

In addition to not having a response to fiber supplementation due to lack of the required microbes, it is also possible that there are specific instances of colorectal cancer where SCFAs may be beneficial. One limitation of current research into the effect of SCFAs in CRC has been that all tumors are treated as the same type. However, there are known differences in the types of mutations that occur [@Fearon1990] and treating all tumors as equal may actually hide any benefit that could be found in specific subsets of individuals. Similar to the idea of specific immunotherapy for specific tumors [@Thomas2016], SCFAs may have beneficial effects for specific types of colorectal tumors. Future research will need to test if this is a valid hypothesis. Regardless of this limitation, our results in combination to previous randomized controlled trials on fiber supplementation suggests that using SCFAs as a general treatment for colorectal cancer is unlikely to provide a reduction in tumorigenesis.

One possible technical limitation is that a fecal sample may not be an ideal type of biospecimen and that the effect SCFAs have on tumorigenesis is only detected in the colon. However, this is unlikely to be a major confounder. First, most *in vivo* studies as well as human studies have used fecal material in their analysis [@Ohigashi2013; @Bishehsari2018]. Second, previous studies that measure SCFA changes after fiber supplementation use fecal material to track these responses with a great deal of success [@scfa_measures_venkataraman2016]. Although there are limitations with the current research on SCFAs and colorectal tumors, technical limitations are less likely to be cause of this. Additionally, as mentioned earlier, our observations along with the randomized controlled trials on fiber supplementation in tumor recurrence [@Yao2017] provide evidence that these specific metabolites may not be protective or used as a general treatment option in colorectal cancer. Yet, taxa that are associated with SCFA production are consistently higher in indivdiuals without colon tumors than patients with carcinomas [@Shah2017; @normalization_Sze2017; @meta_analysis_crc_Sze2018].

The potential protection against colorectal cancers may not be from SCFAs, even though taxa associated with their production are higher in individuals without tumors versus patients with carcinomas [@Shah2017; @normalization_Sze2017; @meta_analysis_crc_Sze2018]. Indeed our data would support the contention that the taxa are similar to those associated with SCFA production but that these specific microbes or OTUs themselves are not associated with SCFAs. In particular, our results showing that different OTUs from the same taxonomic classification are in tumor and SCFA Random Forest models supports this hypothesis. This leads to the possibilty that protection may be through two other routes. First, there could be a different pathway or other less extensively studied metabolites that provides the necessary protection against tumorigenesis. Alternatively, protection may not occur via a metabolite but instead through niche exclusion of mouth-associated microbes (e.g. *Fusobacterium*, *Porphyromonas*, *Parvimonas*, *Peptostreptococcus* [@Zeller2014; @Baxter2016; @Kostic2011]). The idea of niche exclusion is similar to how the community protects against *Clostridium difficile* infection [@Theriot2015] with chronic inflammation replacing the role of antibiotics. Even though we did not find lower concentrations of SCFAs associated with colorectal tumors, we think that there are many exciting new avenues to explore because of these results.

@@

Our observations found no difference in SCFA concentration, their utility as a classification tool, or for genes of enzymes involved in SCFA synthesis between individuals without colon tumors and patients with either adenoma or carcinoma tumors. Although these results are different than other reports in the literature, they do align with the randomized controlled trials that have tested fiber use in preventing colorectal tumor recurrence. Additionally, these observations suggest that resident microbes that are not involved in SCFA production may be the important resident community members involved with preventing tumorigenesis. By focusing on the alternative mechanisms that are associated with these non-SCFA producing resident microbes, the identification of more promising therapeutic options for use in treating colorectal cancer may be found.

### 2018-07-27 submission/manuscript.tex
update discussion and conclusion sections

Much of the previous research on SCFA benefit to human CRC has been
illustrated in model systems (16). Many SCFAs are produced through the
breakdown of fiber (14) and a recent study in mice found that fiber
supplementation increased SCFA concentrations and decreased tumor burden
(18). Additionally, SCFAs such as butyrate can inhibit tumor growth in
\emph{in vitro} experiments (17). Yet, observations in humans has been
mixed. Previous case/control studies found associations with lower SCFA
concentrations in individuals with carcinoma tumors (19). However,
individual randomized controlled trials and a recent meta-analysis on
fiber supplementation to prevent tumor recurrence has found no benefit
(20, 23). Our results align with what has been reported in
randomized-controlled trials, that SCFAs do not provide general
protection against colorectal tumors. It is possible though that there
are specific instances where SCFAs may be beneficial.

One limitation of current research into the effect of SCFAs in CRC has
been that all tumors are treated as the same type. However, there are
known differences in the types of mutations that occur (24) and treating
all tumors as equal may actually hide any benefit that could be found in
specific subsets of individuals. Similar to the idea of specific
immunotherapy for specific tumors (25), SCFAs may have beneficial
effects for specific types of colorectal tumors. Future research will
need to test if this is a valid hypothesis. Another limitation is that a
fecal sample may not be an ideal type of biospecimen and that the effect
SCFAs have on tumorigenesis is only detected in the colon. However, most
\emph{in vivo} studies as well as human studies have used fecal material
in their analysis (18, 19). Additionally, studies that measure SCFA
changes after fiber supplementation use fecal material to track these
responses (26). Although there are limitations with the current research
on SCFAs and colorectal tumors, our observations along with the
randomized controlled trials on fiber supplementation in tumor
recurrence (20) provide evidence that these specific metabolites may not
be protective. Yet, taxa that are associated with SCFA production are
consistently higher in indivdiuals without colon tumors than patients
with carcinomas (10, 11, 27).

The potential protection against colorectal cancers may not be from
SCFAs even though taxa associated with their production are higher in
individuals without tumors versus patients with carcinomas (10, 11, 27).
Protection could be via a different pathway and by extension other
metabolites that have not been extensively studied. Alternatively,
protection may not occur via a metabolite but instead through niche
exclusion of mouth-associated microbes (e.g. \emph{Fusobacterium},
\emph{Porphyromonas}, \emph{Parvimonas}, \emph{Peptostreptococcus} (6,
12, 13)). The idea of niche exclusion is similar to how the community
protects against \emph{Clostridium difficile} infection (28) with
chronic inflammation replacing the role of antibiotics. Although we did
not find lower concentrations of SCFAs associated with colorectal
tumors, we think that there are many exciting new avenues to explore
because of these results.

@@

\textbf{Measuring specific SCFAs.} Our protocol for the measurement of
acetate, butyrate, and propionate followed a previously published
protocol that used a High-Performance Liquid Chromatography (HPLC)
machine (26). The following changes to this protocol included the use of
frozen fecal samples suspended in 1ml of PBS instead of fecal
suspensions in DNA Genotek OmniGut tubes, and the use of the actual
weight of fecal samples instead of the average weight for SCFA
concentration normalizations. These methodological changes did not
affect the overall median concentrations of these SCFAs between the two
studies (see Table 1 (26) and Figure 1 here).

@@

\hypertarget{ref-Fearon1990}{}
24. \textbf{Fearon ER}, \textbf{Vogelstein B}. 1990. A genetic model for
colorectal tumorigenesis. Cell \textbf{61}:759--767.
doi:\href{https://doi.org/10.1016/0092-8674(90)90186-i}{10.1016/0092-8674(90)90186-i}.

\hypertarget{ref-Thomas2016}{}
25. \textbf{Thomas X}, \textbf{Heiblig M}. 2016. The development of
agents targeting the BCR-ABL tyrosine kinase as philadelphia
chromosome-positive acute lymphoblastic leukemia treatment. Expert
Opinion on Drug Discovery \textbf{11}:1061--1070.
doi:\href{https://doi.org/10.1080/17460441.2016.1227318}{10.1080/17460441.2016.1227318}.

\hypertarget{ref-scfa_measures_venkataraman2016}{}
26. \textbf{Venkataraman A}, \textbf{Sieber JR}, \textbf{Schmidt AW},
\textbf{Waldron C}, \textbf{Theis KR}, \textbf{Schmidt TM}. 2016.
Variable responses of human microbiomes to dietary supplementation with
resistant starch. Microbiome \textbf{4}.
doi:\href{https://doi.org/10.1186/s40168-016-0178-x}{10.1186/s40168-016-0178-x}.

>>>>

Although SCFAs have been shown to be anti-tumorigenic, most of these
studies have been performed in model systems (16, 17). Additionally,
many of the \emph{in vivo} studies investigate proxies such as fiber
supplementation rather than SCFAs directly (14). Although it is well
known that breakdown products from gut bacteria results in SCFA
production, fiber effects on tumorigenesis may be through other
mechanisms in these \emph{in vivo} model systems. In addition to using
fiber supplementation as a proxy for SCFA treatment, observations on the
benefit of SCFAs in preventing tumorigenesis has been mixed. In previous
case/control studies lower SCFA concentrations between patients with
carcinomas and those without carcinomas was observed (19). Yet, this is
in contrast to multiple randomized-controlled trials that have found no
difference between patients who do and do not get fiber supplementation
to try and prevent tumor recurrence (20, 23). This could be because
fiber has other actions in humans besides providing an energy source for
bacteria to breakdown to make SCFAs. It could also be due to the fact
that SCFA concentrations and responses to fiber can vary quite a bit
between healthy individuals (24). This information taken together with
our observations would suggest that either indviduals who do not respond
to fiber supplementation would need to acquire these bacteria to achieve
a benefit or that SCFAs provide little to no benefit as an
anti-tumorigenic compound in colorectal cancer.

In addition to not having a response to fiber supplementation due to
lack of the required microbes, it is also possible that there are
specific instances of colorectal cancer where SCFAs may be beneficial.
One limitation of current research into the effect of SCFAs in CRC has
been that all tumors are treated as the same type. However, there are
known differences in the types of mutations that occur (25) and treating
all tumors as equal may actually hide any benefit that could be found in
specific subsets of individuals. Similar to the idea of specific
immunotherapy for specific tumors (26), SCFAs may have beneficial
effects for specific types of colorectal tumors. Future research will
need to test if this is a valid hypothesis. Regardless of this
limitation, our results in combination to previous randomized controlled
trials on fiber supplementation suggests that using SCFAs as a general
treatment for colorectal cancer is unlikely to provide a reduction in
tumorigenesis.

One possible technical limitation is that a fecal sample may not be an
ideal type of biospecimen and that the effect SCFAs have on
tumorigenesis is only detected in the colon. However, this is unlikely
to be a major confounder. First, most \emph{in vivo} studies as well as
human studies have used fecal material in their analysis (18, 19).
Second, previous studies that measure SCFA changes after fiber
supplementation use fecal material to track these responses with a great
deal of success (24). Although there are limitations with the current
research on SCFAs and colorectal tumors, technical limitations are less
likely to be cause of this. Additionally, as mentioned earlier, our
observations along with the randomized controlled trials on fiber
supplementation in tumor recurrence (20) provide evidence that these
specific metabolites may not be protective or used as a general
treatment option in colorectal cancer. Yet, taxa that are associated
with SCFA production are consistently higher in indivdiuals without
colon tumors than patients with carcinomas (10, 11, 27).

The potential protection against colorectal cancers may not be from
SCFAs, even though taxa associated with their production are higher in
individuals without tumors versus patients with carcinomas (10, 11, 27).
Indeed our data would support the contention that the taxa are similar
to those associated with SCFA production but that these specific
microbes or OTUs themselves are not associated with SCFAs. In
particular, our results showing that different OTUs from the same
taxonomic classification are in tumor and SCFA Random Forest models
supports this hypothesis. This leads to the possibilty that protection
may be through two other routes. First, there could be a different
pathway or other less extensively studied metabolites that provides the
necessary protection against tumorigenesis. Alternatively, protection
may not occur via a metabolite but instead through niche exclusion of
mouth-associated microbes (e.g. \emph{Fusobacterium},
\emph{Porphyromonas}, \emph{Parvimonas}, \emph{Peptostreptococcus} (6,
12, 13)). The idea of niche exclusion is similar to how the community
protects against \emph{Clostridium difficile} infection (28) with
chronic inflammation replacing the role of antibiotics. Even though we
did not find lower concentrations of SCFAs associated with colorectal
tumors, we think that there are many exciting new avenues to explore
because of these results.

@@

\textbf{Measuring specific SCFAs.} Our protocol for the measurement of
acetate, butyrate, and propionate followed a previously published
protocol that used a High-Performance Liquid Chromatography (HPLC)
machine (24). The following changes to this protocol included the use of
frozen fecal samples suspended in 1ml of PBS instead of fecal
suspensions in DNA Genotek OmniGut tubes, and the use of the actual
weight of fecal samples instead of the average weight for SCFA
concentration normalizations. These methodological changes did not
affect the overall median concentrations of these SCFAs between the two
studies (see Table 1 (24) and Figure 1 here).

@@

\hypertarget{ref-scfa_measures_venkataraman2016}{}
24. \textbf{Venkataraman A}, \textbf{Sieber JR}, \textbf{Schmidt AW},
\textbf{Waldron C}, \textbf{Theis KR}, \textbf{Schmidt TM}. 2016.
Variable responses of human microbiomes to dietary supplementation with
resistant starch. Microbiome \textbf{4}.
doi:\href{https://doi.org/10.1186/s40168-016-0178-x}{10.1186/s40168-016-0178-x}.

\hypertarget{ref-Fearon1990}{}
25. \textbf{Fearon ER}, \textbf{Vogelstein B}. 1990. A genetic model for
colorectal tumorigenesis. Cell \textbf{61}:759--767.
doi:\href{https://doi.org/10.1016/0092-8674(90)90186-i}{10.1016/0092-8674(90)90186-i}.

\hypertarget{ref-Thomas2016}{}
26. \textbf{Thomas X}, \textbf{Heiblig M}. 2016. The development of
agents targeting the BCR-ABL tyrosine kinase as philadelphia
chromosome-positive acute lymphoblastic leukemia treatment. Expert
Opinion on Drug Discovery \textbf{11}:1061--1070.
doi:\href{https://doi.org/10.1080/17460441.2016.1227318}{10.1080/17460441.2016.1227318}.
